---
title: "Somatic Mutation Analysis with maftools — PAAD-TB Cohort"
subtitle: "Data loading, TMB, oncoplot, mutational signatures, somatic interactions and lolliplots"
author: "Generated with Claude"
date: "`r format(Sys.Date(), '%d/%m/%Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    theme: flatly
    code_folding: show
    fig_width: 10
    fig_height: 7
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center"
)
```

# Introduction

This document reproduces, step by step, a somatic variant analysis workflow
using the Bioconductor package **maftools** (Mayakonda *et al.*, 2018,
*Genome Research*) on the MAF file of the **PAAD-TB** cohort
(`PAAD-TB.final_analysis.maf`).

`maftools` is designed to work directly with MAF files (Mutation
Annotation Format), the standard format used by TCGA and many somatic
variant calling pipelines (e.g., Mutect2/VarScan outputs annotated with
Oncotator or VEP + `vcf2maf`). Each analysis in this document is documented
with references to the official manual and the package tutorial
("vignette"):

- Official vignette: `vignette("maftools")` or
  <https://bioconductor.org/packages/release/bioc/vignettes/maftools/inst/doc/maftools.html>
- Reference manual (functions and parameters):
  <https://bioconductor.org/packages/release/bioc/manuals/maftools/man/maftools.pdf>
- Repository: <https://github.com/PoisonAlien/maftools>

The workflow covers:

1. MAF loading and summary (`read.maf`, `plotmafSummary`).
2. **Tumor Mutational Burden (TMB)** calculation.
3. Most mutated genes (`mafbarplot`, `getGeneSummary`).
4. **Oncoplot** (mutation matrix by gene/sample).
5. **Mutational signature** analysis (tri-nucleotide context,
   NMF extraction, comparison with COSMIC signatures).
6. Signature bar chart (`plotSignatures`).
7. **Cosine similarity** heatmap.
8. **Somatic interaction** heatmap (co-occurrence / mutual exclusivity).
9. **Lolliplot** for `DNMT3A` (transcript `NM_022552`).
10. Analysis of **Pfam domains** affected by the mutations.
11. Additional analyses recommended by the package itself (Ti/Tv,
    tumor heterogeneity, oncogenic pathways, rainfall plots, comparison
    with TCGA, etc.).

---

# Required packages

`maftools` depends on Bioconductor. For the mutational signature analysis
we additionally need a reference genome (`BSgenome`) matching the assembly
used in the MAF (here, GRCh37/hg19, as indicated by `assembly = "37"` in
the original `read.maf`), and `NMF` for the non-negative matrix
factorization underlying `extractSignatures`.

```{r install-packages, eval=FALSE}
# Run only if the packages are not installed.
if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}

BiocManager::install(c(
  "maftools",
  "NMF",
  "BSgenome.Hsapiens.UCSC.hg19",   # GRCh37/hg19 reference genome
  "pheatmap",
  "RColorBrewer"
))
```

```{r libraries}
library(maftools)
library(NMF)
library(BSgenome.Hsapiens.UCSC.hg19)
library(pheatmap)
library(RColorBrewer)
library(R.utils)
```

---

# Data loading

## Directories and input file

We replicate the indicated path configuration, adapted to R syntax
(R has no `os.makedirs`/`Path`; we use `dir.create` and `file.path`).

```{r paths}
# Results directory (equivalent to os.makedirs("results_paad", exist_ok=TRUE))
results_dir <- "results_laml_maftools"
dir.create(results_dir, showWarnings = FALSE, recursive = TRUE)

# Working directory (equivalent to WORKING_DIR = Path(...))
WORKING_DIR <- "/home/albalopez/Documentos/PyMut_DocsAlba"

# Path to the MAF
maf_path <- file.path(WORKING_DIR, "data", "tcga_laml.maf.gz") #tcga_laml.maf.gz #PAAD-TB.final_analysis.maf
maf_path 
```

## Reading the MAF with `read.maf`

`read.maf()` is the entry point of the whole package: it validates the
file, computes summary statistics per sample and per gene, classifies
variants into silent/non-silent according to `vc_nonSyn`, and returns an
S4 object of class `MAF` on which all other functions operate
(`plotmafSummary`, `oncoplot`, `tmb`, `lollipopPlot`, etc.). See
`?read.maf` for the full parameter details.

Note: in R the correct function name is `read.maf` (with a dot), not
`read_maf`. The genome assembly is not itself a parameter of
`read.maf`; we will take it into account later when choosing the
`BSgenome` (GRCh37/hg19) for the mutational signature analysis.

```{r read-maf}
mutations <- read.maf(
  maf                       = maf_path,
  removeDuplicatedVariants  = TRUE,   # maftools default behaviour
  isTCGA                    = FALSE   # the barcode does not follow the 12+ character TCGA format
)

mutations
```

```{r maf-basic-info}
# Quick summaries included in the MAF object
getSampleSummary(mutations)
getGeneSummary(mutations)
getClinicalData(mutations)
# getFields(mutations)
```

---

# Summary Plot

`plotmafSummary()` generates a composite figure with: (1) variant
classification, (2) variant type, (3) SNV class, (4) variants per sample
(stacked barplot), (5) boxplot of variants per classification, and (6) the
most mutated genes. It is the recommended first "quality control" look
from the official tutorial before any finer-grained analysis.

```{r summary-plot, fig.height=8}
plotmafSummary(
  maf        = mutations,
  rmOutlier  = TRUE,
  addStat    = "median",
  dashboard  = TRUE,
  titvRaw    = FALSE
)
```

---

# TMB (Tumor Mutational Burden) calculation

`tmb()` computes the mutational burden (non-silent mutations per megabase)
for each sample, using maftools' default reference capture exome size
(`captureSize = 50` Mb, adjustable to the actual sequencing kit if known).
It is a key metric in precision oncology (associated with response to
immunotherapy).

```{r tmb}
tmb_table <- tmb(
  maf         = mutations,
  captureSize = 50,   # Mb — adjust to the real panel/exome size if known
  logScale    = TRUE
)

tmb_table
write.csv(tmb_table, file.path(results_dir, "TMB_por_muestra.csv"), row.names = FALSE)
```
```{r}
compute_stats <- function(x) {
  c(
    Count  = length(x),
    Median = median(x, na.rm = TRUE),
    Mean   = mean(x, na.rm = TRUE),
    Min    = min(x, na.rm = TRUE),
    Max    = max(x, na.rm = TRUE),
    Q1     = quantile(x, 0.25, na.rm = TRUE),
    Q3     = quantile(x, 0.75, na.rm = TRUE),
    Std    = sd(x, na.rm = TRUE)
  )
}

TMB_statistics <- rbind(
  data.frame(Metric = "total", t(compute_stats(tmb_table$total))),
  data.frame(Metric = "total_perMB", t(compute_stats(tmb_table$total_perMB)))
)

rownames(TMB_statistics) <- NULL
TMB_statistics
```
```{r}
write.csv(
  TMB_statistics,
  file = file.path(results_dir, "TMB_statistics.csv"),
  row.names = FALSE
)
```

The resulting plot shows, per sample, the TMB (log10 mutations/Mb) compared
against the TMB distribution of TCGA cohorts, making it possible to
contextualize whether the PAAD-TB cohort is hyper/hypomutated relative to
other tumor types.

---

# Top mutated genes

## `mafbarplot`: ranking of most mutated genes

`mafbarplot()` produces a horizontal barplot with the N most frequently
mutated genes (colored by variant type by default), a more compact
alternative to the "top genes" section of `plotmafSummary`.

```{r top-genes-barplot}
mafbarplot(maf = mutations, n = 20, fontSize = 0.6, legendfontSize = 0.8)
```

## Table of most mutated genes

```{r top-genes-table}
gene_summary <- getGeneSummary(mutations)
head(gene_summary[order(-MutatedSamples)], 20)
write.csv(gene_summary, file.path(results_dir, "gene_summary.csv"), row.names = FALSE)
```

---

# Oncoplot

`oncoplot()` is maftools' flagship function: it draws a matrix of genes
(rows) x samples (columns) colored by consequence type, with marginal bars
of mutation frequency per gene and per sample, and optional clinical
annotations. It is the R equivalent of the "waterfall plot"/OncoPrinter in
cBioPortal.

```{r oncoplot, fig.height=8}
oncoplot(
  maf          = mutations,
  top          = 20,          # top 20 most mutated genes
  removeNonMutated = TRUE,
  showTumorSampleBarcodes = FALSE,
  draw_titv    = TRUE         # adds Ti/Tv panel
)
```

If the `mutations` object included clinical data (`clinicalData` in
`read.maf`), column annotations could be added with the argument
`clinicalFeatures` and the oncoplot sorted with `sortByAnnotation = TRUE`.

---

# Mutational signature analysis

The mutational signature analysis follows the standard workflow described
in the maftools vignette *"Mutational Signatures"*:

1. `trinucleotideMatrix()`: extracts, for each SNP mutation, the
   tri-nucleotide context (96 possible combinations) using the reference
   genome. Here we use `BSgenome.Hsapiens.UCSC.hg19`, consistent with the
   GRCh37 assembly indicated in the original MAF loading (`assembly="37"`).
2. `estimateSignatures()`: runs NMF (non-negative matrix factorization)
   for several values of *k* (number of signatures) and computes the
   cophenetic correlation, which helps decide the optimal number of
   signatures.
3. `plotCophenetic()`: visualizes that metric to choose *k*.
4. `extractSignatures()`: extracts de novo signatures for the chosen *k*.
5. `compareSignatures()`: compares the extracted signatures against the
   COSMIC reference signatures (v2 "legacy" or v3 SBS) by cosine
   similarity, and suggests the best biological match (e.g., aging,
   tobacco, mismatch-repair deficiency signatures, etc.).

```{r trinucleotide-matrix}
maf_tnm <- trinucleotideMatrix(
  maf        = mutations,
  ref_genome = "BSgenome.Hsapiens.UCSC.hg19",
  prefix     = "chr",     # the UCSC genome uses the "chr" prefix; set to "" if the MAF already uses it
  add        = TRUE,
  useSyn     = TRUE
)
```

```{r apobec-enrichment}
# trinucleotideMatrix also reports the enrichment of the APOBEC signature,
# of interest in many solid tumors.
maf_tnm$APOBEC_scores
```

## Estimating the optimal number of signatures

```{r estimate-signatures, results='hide'}
sig_estimate <- estimateSignatures(
  mat        = maf_tnm,
  nTry       = 6,     # tries k = 1..6 signatures
  pConstant  = 0.1
)
```

```{r cophenetic-plot}
plotCophenetic(res = sig_estimate)
```

The cophenetic correlation plot shows, for each *k*, the stability of the
NMF solution; the usual criterion (official tutorial) is to choose the
*k* just before the first notable drop in the correlation.

## Extraction of de novo signatures

```{r extract-signatures, results='hide'}
# Adjust n_sig to the optimal value observed in plotCophenetic (example: 3)
n_sig <- 3

sig_extract <- extractSignatures(
  mat   = maf_tnm,
  n     = n_sig,
  pConstant = 0.1
)
```

## Comparison with COSMIC signatures (cosine similarity)

```{r compare-signatures-v2}
# Comparison against COSMIC v2 legacy signatures (30 "classic" signatures)
sig_compare_v2 <- compareSignatures(
  nmfRes = sig_extract,
  sig_db = "legacy"
)

sig_compare_v2$best_match
```

```{r compare-signatures-v3}
# Comparison against COSMIC v3 signatures (SBS, more recent and granular)
sig_compare_v3 <- compareSignatures(
  nmfRes = sig_extract,
  sig_db = "SBS"
)

sig_compare_v3$best_match
```

---

# Signature Bar Chart

`plotSignatures()` draws, for each extracted signature, the 96
tri-nucleotide context profile (bars colored by the 6 substitution
classes: C>A, C>G, C>T, T>A, T>C, T>G), together with the name of the most
similar COSMIC signature and its cosine similarity.

```{r signature-barchart, fig.height=8}
plot_signatures <- plotSignatures(
  nmfRes = sig_extract,
  title_size = 1.0,
  sig_db = "SBS"
)
```
```{r}
# Output path inside the results folder
png(file = file.path(results_dir, "signatures_SBS.png"), width = 2300, height = 1500, res = 300)
plotSignatures(
  nmfRes = sig_extract,
  title_size = 1.4,
  font_size = 1.2,
  sig_db = "SBS"
)
dev.off()
```

---

# Cosine Similarity Heatmap

Besides the `best_match` returned by `compareSignatures()`, it is very
informative to visualize the full cosine similarity matrix between the
extracted signatures and all reference signatures, to detect secondary
matches or composite signatures.

```{r cosine-heatmap, fig.height=7}
cosine_matrix <- sig_compare_v3$cosine_similarities

pheatmap(
  cosine_matrix,
  cluster_rows       = TRUE,
  cluster_cols       = TRUE,
  color              = colorRampPalette(rev(brewer.pal(9, "RdBu")))(100),
  main               = "Cosine similarity: de novo signatures vs. COSMIC SBS",
  fontsize_row       = 10,
  fontsize_col       = 6
)
```

`maftools` also offers an equivalent built-in shortcut:

```{r cosine-heatmap-builtin, fig.height=6}
compareSignatures(
  nmfRes  = sig_extract,
  sig_db  = "SBS"
  #input_sig_db = NULL
)
```

---

# Somatic interactions heatmap

`somaticInteractions()` evaluates, for each pair of genes among the N most
mutated, whether their co-mutation in the same sample occurs more often
than expected by chance (co-occurrence) or less often (mutual exclusivity),
using Fisher's exact test (or `pairwise.fisher.test`), and represents it
as a triangular heatmap with significance asterisks.

```{r somatic-interactions, fig.height=8}
somatic_int <- somaticInteractions(
  maf        = mutations,
  top        = 20,
  pvalue     = c(0.05, 0.01)
)

head(somatic_int)
write.csv(somatic_int, file.path(results_dir, "somatic_interactions.csv"), row.names = FALSE)
```

---

# Lolliplot — DNMT3A

`lollipopPlot()` draws the mutations of a specific gene over its protein
structure, marking the position of each amino-acid change as a "lollipop"
whose height indicates its frequency. It requires the MAF to have a column
with the amino-acid change (`aa_col`) and, optionally, the reference
transcript can be fixed (`refSeqID`) to ensure that domain coordinates
match the correct isoform.

```{r lollipop-dnmt3a, fig.height=6}
lollipopPlot(
  maf          = mutations,
  gene         = "DNMT3A",
  AACol        = "Protein_Change",
  refSeqID     = "NM_022552"
#   showMutationRate = TRUE,
#   labelPos     = 882,
#   pointSize = 2
 )
```

```{r}
png(
  filename = "dnmt3a_lollipop_300dpi.png",
  units = "in",
  width = 16,
  height = 8,
  res = 300
)

op <- par(oma = c(0, 0, 3, 0), mar = c(5, 4, 5, 2) + 0.1)

lollipopPlot(
  maf = mutations,
  gene = "DNMT3A",
  AACol = "Protein_Change",
  labelPos = 882,
  legendTxtSize = 1.4,
  domainLabelSize = 1.1,
  axisTextSize = c(1.2, 1.2),
  titleSize = c(1.8, 1.2),
  labPosSize = 1.4,
  pointSize = 3.0,
  showLegend = TRUE
)

par(op)
dev.off()
```

If `refSeqID = "NM_022552"` does not find an exact match in maftools'
internal Pfam domain database, the function automatically falls back to
the canonical transcript available; it is advisable to check the console
message to confirm which isoform was finally used.

---

# Pfam domain analysis

`pfamDomains()` combines the amino-acid change annotation of the whole MAF
with maftools' integrated Pfam domain database, indicating which protein
domains concentrate the most mutations across the entire cohort (not just
one gene), very useful for prioritizing functionally relevant domains
(e.g., catalytic domains, DNA-binding domains, etc.).

```{r pfam-domains, fig.height=7}
pfam_result <- pfamDomains(
  maf     = mutations,
  AACol   = "Protein_Change",
  top     = 10
)

pfam_result$proteinSummary
pfam_result$domainSummary
write.csv(pfam_result$domainSummary, file.path(results_dir, "pfam_domain_summary.csv"), row.names = FALSE)
```

---

# Additional analyses recommended by maftools

These analyses were not explicitly requested, but they are a standard part
of the maftools tutorial and take advantage of the data already loaded in
`mutations`, providing additional context for the biological
interpretation.

## Transitions / Transversions (Ti/Tv)

```{r titv}
titv_result <- titv(maf = mutations, plot = FALSE, useSyn = TRUE)
plotTiTv(res = titv_result)
```

Classifies each SNV as a transition (purine<->purine or
pyrimidine<->pyrimidine) or a transversion, and summarizes the proportion
per sample; characteristic patterns (e.g., excess of C>T) may suggest
specific mutagenic processes (UV light, tobacco, aging, 5-methylcytosine
deamination).

## Rainfall plot (localized hypermutation / kataegis)

Useful if a specific sample has many mutations; it identifies genomic
regions with localized hypermutation events ("kataegis"), frequently
associated with APOBEC activity.

```{r rainfall, eval=FALSE}
# Run on a specific sample of interest, replacing "SAMPLE_ID"
rainfallPlot(
  maf        = mutations,
  tsb        = "SAMPLE_ID",
  detectChangePoints = TRUE,
  pointSize  = 0.4
)
```

## Comparison with TCGA cohorts

`tcgaCompare()` places the TMB of the PAAD-TB cohort in the context of all
TCGA cohorts, generating a comparative boxplot sorted by TMB median — a
natural extension of the TMB calculation already performed.

```{r tcga-compare, fig.height=7}
tcga_compare <- tcgaCompare(
  maf        = mutations,
  cohortName = "PAAD-TB",
  logscale   = TRUE,
  capture_size = 50
)
```

## Oncogenic pathways (OncoKB / canonical pathways)

`OncogenicPathways()` summarizes the percentage of altered samples in each
of the 10 oncogenic signaling pathways curated by Sanchez-Vega
*et al.* 2018 (RTK-RAS, PI3K, WNT, TP53, etc.), and
`PlotOncogenicPathways()` details, for a specific pathway, which genes
contribute and with what exclusivity pattern.

```{r oncogenic-pathways, fig.height=6}
oncogenic_pathways <- OncogenicPathways(maf = mutations)
oncogenic_pathways
```

```{r plot-oncogenic-pathway-example, fig.height=6}
# Example with the RTK-RAS pathway, usually relevant in PAAD (KRAS)
PlotOncogenicPathways(maf = mutations, pathways = "RTK-RAS")
```

## Tumor heterogeneity / clonality (VAF-based)

`inferHeterogeneity()` estimates tumor heterogeneity clusters from the
variant allele frequency (VAF), useful for exploring clonal/subclonal
evolution; it requires allelic depth columns in the MAF (`t_depth`,
`t_alt_count` or equivalents).

```{r heterogeneity, fig.height=6}
heterogeneity <- inferHeterogeneity(
  maf     = mutations,
  vafCol  = "t_AF"   # adjust to the real VAF column name in the MAF, if it exists
)
plotClusters(clusters = heterogeneity)
```

## VAF distribution by gene

```{r plot-vaf, fig.height=6}
plotVaf(maf = mutations, vafCol = "t_AF", top = 15)
```

## Potentially "driver" genes (dN/dS with `oncodrive`)

`oncodrive()` implements the OncodriveCLUST algorithm, which identifies
genes whose mutations cluster at specific protein positions more than
expected by chance, an indicator of positive selection typical of driver
genes.

```{r oncodrive, fig.height=6}
oncodrive_result <- oncodrive(
  maf         = mutations,
  AACol       = "Protein_Change",
  minMut      = 5,
  pvalMethod  = "zscore"
)
plotOncodrive(res = oncodrive_result, fdrCutOff = 0.1, useFraction = TRUE)
```

---

# Saving results

```{r save-maf-object}
saveRDS(mutations, file.path(results_dir, "maf_object.rds"))
```

All csv/rds files generated in this document are stored in the
`results_paad/` folder, analogous to the one created with
`os.makedirs("results_paad", exist_ok=True)` in the original snippet.

---

# References

- Mayakonda, A., Lin, D.C., Assenov, Y., Plass, C., Koeffler, H.P. (2018).
  *Maftools: efficient and comprehensive analysis of somatic variants in
  cancer*. Genome Research, 28:1747-1756.
- Official maftools vignette:
  <https://bioconductor.org/packages/release/bioc/vignettes/maftools/inst/doc/maftools.html>
- Package page on Bioconductor:
  <https://bioconductor.org/packages/release/bioc/html/maftools.html>
- GitHub repository (issues, additional examples):
  <https://github.com/PoisonAlien/maftools>
- Sanchez-Vega, F. *et al.* (2018). *Oncogenic Signaling Pathways in The
  Cancer Genome Atlas*. Cell, 173(2), 321-337 (basis of `OncogenicPathways`).

```{r session-info}
sessionInfo()
```
